Comparing Exchangeable Effects (EE) and Exchangeable Standardized Effects (ET) models

### 2016-07-24 analysis/CompareEEvsETmodel.Rmd
update from model parameter to alpha parameter

A key modelling assumption is that the effects $\beta_j$ are identically distributed
from a unimodal distribution $g$. 
In particular, ashr assumes that $\beta_j$ is independent of the standard error $s_j$.
That is,
$$\beta_j | s_j \sim g(\cdot)     \quad (*).$$

Here we consider an alternative modelling assumption:
$$\beta_j/s_j | s_j \sim g(\cdot)   \quad (**).$$
Under this alternative assumption
the *t statistics* $\betahat_j/s_j$ are identically distributed from a unimodal
distribution. Under this assumption the expected size of the *unstandardized* effect $\beta_j$ depends on the standard error $s_j$: the larger $s_j$ is, the larger (in absolute value) $\beta_j$ is expected to be.

@@

It is straightforward to use the ashr package to perform analysis under the ET model:
simply specify model="ET". (Internally, this causes ash to 
replace betahat with the standardized betahat, $\hat\beta_j/s_j$, and the standard errors for these standarized betahat values with 1; there there is some bookkeeping to be done to make sure we return the right likelihoods and posteriors for the original beta, and not for these standardized values... ash takes care of this.) It is also straightforward to compare
the two competing modelling assumptions (EE vs ET) by computing the log likelihood ratio, $log[p_{EE}(\hat\beta | s, \hat g_{EE})/p_{ET}(\hat\beta | s, \hat g_{ET})]$.

@@

Here is the EE analysis.
```{r}
  library(ashr)
  ashEE.res = ash(betahat, betahat.se,method="fdr",model="EE")
```

And here is how we can perform the ET analysis,
```{r}
  ashET.res = ash(betahat, betahat.se,method="fdr", model="ET")
```

@@

The above illustrates these ideas on simulations from the ET model. For comparison,
we now provide results for simulations under the EE model.
```{r}
  set.seed(1234)
  nsamp=1000
  betahat.se = rgamma(nsamp,1,1)
  beta = rnorm(nsamp) #simulate effects under the original assumption 
  betahat = rnorm(nsamp,beta,betahat.se)
  zscore = betahat/betahat.se
  pval = pchisq(zscore^2,df=1,lower.tail=FALSE)
  ashEE.res = ash(betahat, betahat.se,method="fdr",model="EE")
  ashET.res = ash(betahat, betahat.se, method="fdr",model="ET")
  ashEE.res$loglik-ashET.res$loglik
```
So here the log likelihood ratio is positive, indicating that the EE model is preferred
(which is expected since the data were generated under that model).

>>>>

The default model is that the effects $\beta_j$ are identically distributed
from a unimodal distribution $g$. 
In particular, ashr assumes that $\beta_j$ is independent of the standard error $s_j$.
That is,
$$\beta_j | s_j \sim g(\cdot)     \quad (*).$$

Here we consider the more general model:
$$\beta_j/(s_j)^\alpha | s_j \sim g(\cdot)   \quad (**).$$
The case $\alpha=0$ is the model (*) above. The case $\alpha=1$ assumes that 
the *t statistics* $\betahat_j/s_j$ are identically distributed from a unimodal
distribution. Under this assumption the expected size of the *unstandardized* effect $\beta_j$ depends on the standard error $s_j$: the larger $s_j$ is, the larger (in absolute value) $\beta_j$ is expected to be.

@@

It is straightforward to use the ashr package to perform analysis under the ET model:
simply specify alpha=1. (Internally, this causes ash to 
replace betahat with the standardized betahat, $\hat\beta_j/s_j$, and the standard errors for these standarized betahat values with 1; there there is some bookkeeping to be done to make sure we return the right likelihoods and posteriors for the original beta, and not for these standardized values... ash takes care of this.) It is also straightforward to compare
the two competing modelling assumptions (EE vs ET) by computing the log likelihood ratio, $log[p_{EE}(\hat\beta | s, \hat g_{EE})/p_{ET}(\hat\beta | s, \hat g_{ET})]$.

@@

Here is the EE analysis.
```{r}
  library(ashr)
  ashEE.res = ash(betahat, betahat.se,method="fdr",alpha=0)
```

And here is how we can perform the ET analysis,
```{r}
  ashET.res = ash(betahat, betahat.se,method="fdr", alpha=1)
```

@@

The above illustrates these ideas on simulations from the ET model. For comparison,
we now provide results for simulations under the EE model.
```{r}
  set.seed(1234)
  nsamp=1000
  betahat.se = rgamma(nsamp,1,1)
  beta = rnorm(nsamp) #simulate effects under the original assumption 
  betahat = rnorm(nsamp,beta,betahat.se)
  zscore = betahat/betahat.se
  pval = pchisq(zscore^2,df=1,lower.tail=FALSE)
  ashEE.res = ash(betahat, betahat.se,method="fdr",alpha=0)
  ashET.res = ash(betahat, betahat.se, method="fdr",alpha=1)
  ashEE.res$loglik-ashET.res$loglik
```
So here the log likelihood ratio is positive, indicating that the EE model is preferred
(which is expected since the data were generated under that model).
